---
title: "Introduction to chemfuse"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to chemfuse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  eval = FALSE  # do not run by default (requires Python + chemfuse)
)
library(chemfuse)
```

## Overview

The **chemfuse** R package provides an R interface to the Python [chemfuse](https://github.com/hurlab/ChemFuse) package, enabling R users to:

- Search chemical compounds across PubChem, ChEMBL, UniChem, BindingDB, Open Targets, and SureChEMBL
- Retrieve detailed compound data with optional enrichment (ADMET, drug-likeness, binding)
- Perform batch screening of compound libraries
- Map compound identifiers across databases (cross-referencing)
- Compute molecular descriptors and predict ADMET properties locally
- Export results to CSV or Excel

## Installation

### Python Prerequisites

The R package requires the Python `chemfuse` package to be installed:

```bash
pip install chemfuse[all]
```

For ADMET predictions, include the `admet` extra:

```bash
pip install "chemfuse[admet]"
```

### R Package Installation

```r
# From R-universe (recommended)
install.packages("chemfuse", repos = "https://hurlab.r-universe.dev")

# Or from GitHub
# install.packages("remotes")
# remotes::install_github("hurlab/ChemFuse/r-package/chemfuse")
```

### Configure reticulate

If you have multiple Python installations, tell reticulate which one has chemfuse:

```r
library(reticulate)
use_python("/path/to/python")  # or use_virtualenv(), use_condaenv()
```

## Basic Usage

### Searching Compounds

```{r search}
library(chemfuse)

# Search by name
results <- cf_search("aspirin")
print(results)

# Search with multiple databases
results_multi <- cf_search("aspirin", sources = c("pubchem", "chembl"))

# Search by SMILES
results_smiles <- cf_search(
  "CC(=O)Oc1ccccc1C(=O)O",
  query_type = "smiles"
)
```

The `cf_search()` function returns a **tibble** with columns including `cid`, `name`, `smiles`, `inchikey`, `formula`, and `molecular_weight`.

### Retrieving a Single Compound

```{r get}
# Basic retrieval
compound <- cf_get("aspirin")
compound$compound  # tibble with compound data

# With ADMET predictions
enriched <- cf_get("aspirin", admet = TRUE)
enriched$admet  # tibble with ADMET predictions

# Full enrichment
full <- cf_get("aspirin", admet = TRUE, druglikeness = TRUE)
full$druglikeness  # list of drug-likeness filter results
```

### Batch Screening

```{r screen}
# Screen a data frame
library <- data.frame(
  smiles = c(
    "CC(=O)Oc1ccccc1C(=O)O",  # aspirin
    "Cn1cnc2c1c(=O)n(c(=O)n2C)C",  # caffeine
    "c1ccc2c(c1)cc1ccc3cccc4ccc2c1c34"  # pyrene
  ),
  name = c("aspirin", "caffeine", "pyrene")
)

results <- cf_screen(library, sources = c("pubchem"))
print(results)

# Screen from a CSV file
results_csv <- cf_screen("my_compounds.csv")
```

### Cross-Database Identifier Mapping

```{r xref}
# Map by PubChem CID (aspirin = 2244)
xref <- cf_xref(cid = 2244)
print(xref)

# Map by ChEMBL ID
xref_chembl <- cf_xref(chembl_id = "CHEMBL25")

# Map by SMILES
xref_smiles <- cf_xref(smiles = "CC(=O)Oc1ccccc1C(=O)O")
```

The result is a tibble with `database` and `identifier` columns showing all available cross-references.

### Computing Molecular Descriptors

```{r descriptors}
# Compute RDKit descriptors for a compound
desc <- cf_descriptors("CC(=O)Oc1ccccc1C(=O)O")
print(names(desc))  # shows available descriptor columns

# Multiple compounds
descs <- cf_descriptors(c("CCO", "CCC", "CC(=O)O"))
```

### Predicting ADMET Properties

```{r admet}
# Predict ADMET properties
admet <- cf_admet("CC(=O)Oc1ccccc1C(=O)O")
print(admet)

# Check drug-likeness
dl <- cf_druglikeness("CC(=O)Oc1ccccc1C(=O)O")
cat("Passes Lipinski Ro5:", dl$lipinski$passes, "\n")
```

## Exporting Results

```{r export}
# Search for compounds
results <- cf_search("aspirin")

# Export to CSV
cf_to_csv(results, "aspirin_results.csv")

# Export to Excel (requires openxlsx package)
cf_to_excel(results, "aspirin_results.xlsx", sheet = "aspirin")
```

## A Complete Workflow

Here is a complete drug screening workflow:

```{r workflow}
library(chemfuse)
library(dplyr)

# Step 1: Define your compound library
compounds <- data.frame(
  smiles = c(
    "CC(=O)Oc1ccccc1C(=O)O",
    "Cn1cnc2c1c(=O)n(c(=O)n2C)C",
    "OC[C@H]1OC(O)[C@H](O)[C@@H](O)[C@@H]1O"
  ),
  name = c("aspirin", "caffeine", "glucose")
)

# Step 2: Batch screen across PubChem
results <- cf_screen(compounds, sources = c("pubchem"))

# Step 3: Add ADMET predictions for drug candidates
admet_results <- cf_admet(compounds$smiles)

# Step 4: Compute molecular descriptors
descriptors <- cf_descriptors(compounds$smiles)

# Step 5: Cross-reference with other databases
aspirin_xref <- cf_xref(smiles = "CC(=O)Oc1ccccc1C(=O)O")

# Step 6: Export final results
cf_to_csv(results, "screening_results.csv")
```

## Getting Help

- Package documentation: `?cf_search`, `?cf_get`, etc.
- GitHub repository: <https://github.com/hurlab/ChemFuse>
- Python package docs: <https://chemfuse.readthedocs.io>
- Report issues: <https://github.com/hurlab/ChemFuse/issues>
